Social Attention for Autonomous Decision-Making in Dense Traffic

### 2019-09-10 main.tex
Update architecture

% Custom packages
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{xspace}
\usepackage{float}
\usetikzlibrary{arrows,automata}
\usepackage{pgfplots}
\usetikzlibrary{intersections}
\usepackage{subcaption}
\usepackage[flushleft]{threeparttable}
\usepackage{cases}
\usepackage{multirow}

@@

This observation has led the community to turn to learning-based methods, which bear the promise of leveraging data to automatically learn a complex driving policy. In the imitation learning approach, a policy can be trained in a supervised manner to imitate human driving decisions \citep{Pomerleau1989, Ross2011, Bojarski2016, Xu2016, Eraqi2017, Codevilla2017, Rehder2017c, Rezagholiradeh2018, Bansal2018}. Because the cost of human driving data collection at large scale can be prohibitive, another promising approach is train a policy in simulation using reinforcement learning \citep{Cardamone2009, Ross2011}.

@@

Our contributions are the following: first, we propose an attention-based architecture for decision-making involving social interactions, that deals with the varying-size and that accounts for two-way interactions between the ego-vehicle and every other traffic participant
- We evaluate our model on a challenging intersection-crossing task involving up to 15 vehicles perceived simultaneously.

@@

In order to apply a reinforcement learning algorithm such as DQN to an autonomous driving problem, a state space $S$ must first be chosen, that is, a representation of the scene. A vehicle driving on a road can be described in the most general way by it's continuous position, heading and velocity. The composite state (or joint-state) of a road traffic with one ego-vehicle (denoted $s_0$) and N other vehicles can then be described by the set of the states of all vehicles.

\begin{equation*}
s = \left\lbrace s_k \right\rbrace_{k \in [0, N]}\qquad
\text{where}\qquad
s_k = \begin{bmatrix}
x_k & y_k & v^x_k & v^y_k & \cos\psi_k & \sin \psi_k
\end{bmatrix}^T
\end{equation*}

@@

This representation, that we call \emph{list of features}, is illustrated in \autoref{fig:representation} and was used e.g. in \citep{Bai2015, Gindele2015, Song2016, Sunberg2017, Paxton2017, Galceran2017, Chen2017}.

@@

A popular way to address this limitations is to use a spatial matrix representation. Instead of explicitly representing spatial information as variables $x, y$ along with other features $f$ directly inside the state $\{s_k=(x_k,y_k,f_k)\}_{k\in[0,N]}$ indexed on the vehicles, they are instead represented implicitly through the layout of several feature variables $f_ij$ organized in a tensor structure, where the $i,j$ indexes refer to the cells of a quantization of the 2D-space. Note that the size of this tensor is related to the area covered divided by the length of the quantization, which reflects a trade-off between accuracy and computational load.
In an occupancy grid, the $f$ features contains presence information (0-1) and additional channels such as velocity and heading, as in \citep{Isele2017, Fridman2018, Bansal2018, Rehder2017c}. Another example is the use of top-view RGB images \citep{Bagnell2010, Rehder2017, Rehder2017c, Liu2018}.

This permutation invariance property can also be implemented within the policy architecture. A general technique to achieve this is to treat each entity similarly in the early stages -- e.g. through weight sharing -- before reducing them with a projection operator that is itself invariant to permutations, for instance a max-pooling as in \citep{Chen2017} or a sum as in \citep{Qi2016}. A particular instantiation of this idea is the use of attention mechanisms.

\paragraph{Attention mechanisms} {The attention architecture was introduced to enable neural networks to discover inter-dependencies within a variable number of inputs.
It has been used for pedestrian trajectory forecasting in~\cite{Vemula2018} with spatiotemporal graphs and in~\cite{Sadeghian2019CVPR} with spatial and social attention using a generative neural network. In~\cite{Sadeghian2018ECCV}, attention over top-view road scene images for car trajectory forecasting is used. Multi-head attention mechanism has been developed in~\cite{Vaswani2017} for sentence translation. In~\cite{Messaoud2019} a mechanism called non-local multi-head attention is developed. However, this is a spatial attention that does not allow vehicle-to-vehicle attention. In the present work, we use a multi-head social attention mechanism...}

@@

Motivation: 
\begin{enumerate}
	\item Out of a complex scene description, the model should be able to filter information and consider only what is relevant for decision. In other words, the agent should \emph{pay attention} to vehicles that are close or conflict with the planned route;
	\item The model should be able to admit a variable number of vehicles as inputs;
	\item The model should be invariant to the ordering of vehicles.
\end{enumerate}

\begin{figure}[ht]
	\centering
	\begin{tikzpicture}
	\node(X1){ego};
	\node[below of=X1, node distance=0.7cm](X2){vehicle$_{1}$};
	\node[below of=X2, node distance=0.6cm](X3){$\vdots$};
	\node[below of=X3, node distance=0.7cm](X4){vehicle$_{N}$};

@@

	\node[draw, rectangle, right of=ENC1, node distance=1.8cm, below=-0.2cm](TRANS1){\rotatebox{90}{ Ego-attention }};

@@

	\draw (DEC1.east) -- (Y1.west);
	\end{tikzpicture}
	\caption{Block diagram of our model architecture.}
	\label{sch_whole_model}

@@

The multi-head self-attention layers allow vehicle interactions while keeping independence from their number
and ordering.
This mechanism is described in~\cite{Vaswani2017} where it is applied on sentence translation.
In this section we explain its use for vehicle interactions.
The computations made by each attention head is represented on figure~\ref{fig:ego-attention} and are detailed below.

\begin{figure}[ht]
	\centering
	\begin{tikzpicture}[scale=1, every node/.style={scale=1}]
	\node(X1){ego encoding};
	\node[below of=X1, node distance=2cm](X2){$\vdots$};
	\node[below of=X2, node distance=2cm](X3){vehicle$_{N}$ encoding};

@@

%Descriptive features, names keys $k_i$
%The similarity of the query $q_i$ to the keys $k_j$ is assessed through a similarity measure implemented as a scalar product: $<q_i, k_j>$, $QK^T$.
%These similarities are then normalized by a softmax $\sigma$ function and a division by $\sqrt{d_k}}$ where $d_k$ is the dimension of keys and queries. The final value is obtained.

Self attention

Each vehicle should pay attention to specific features from a selection of the other vehicles.
This is made with three steps: ...
Each head produces a different selection of features using a linear projection of
the input tensor resulting in the value tensor $V$.
To identify these features, a key tensor $K$ is associated to each value.
Then, each vehicle (TODO: or just the ego) must select which other vehicle to pay attention to.
For that purpose a query $Q$ is produced to find a selection of keys.
The match score between a key and a query is the dot product is scaled with the square root of the key
dimension $\sqrt{d_k}$ and normalized with a softmax.
This produces an attention matrix that contains coefficients close to $1$ for matching queries and keys
and close to $0$ otherwise.
Finally, this matrix is used to gather the values from $V$.
Thus, the self-attention computation for each head is written:
\begin{equation}
\text{output}=\underbrace{\sigma\left(\frac{QK^T}{\sqrt{d_k}}\right)}_{\text{attention matrix}}V
\label{eq_selfattention}
\end{equation}
The outputs from all heads are concatenated and combined with a linear layer.
The resulting tensor is then added to the input as in residual networks.

Ego attention: only the ego-vehicle produces a query, we get one output feature. We can easily see that the result is permutation invariant: indeed, a permutation $\tau$ will change the order of the rows of keys $K$ and values $V$ but will keep their correspondence. The final result is a sum of values weighted by their key-similarities, which is independent of the ordering. %$\sigma\left(Q\tau(K)^T/\sqrt{d_k}\right)\tau(V) = \tau\left(\sigma(QK^T/\sqrt{d_k})\right)\tau(V) = \sigma\left(QK^T/\sqrt{d_k}\right)V$

@@

\begin{table}[tp]
	\centering
	\begin{threeparttable}
		\caption{Characteristics of the agents}
		\label{tab:agents}
		\begin{tabular}{lccc}
			\toprule
			Architecture & \MLPC & \MLPG & \EgoAtt \\
			\midrule 
			Input sizes & [15, 7] & [?, ?, 7] & [1-15, 7] \\
			Layers sizes & [128, 128] &  [64, 64] & Encoder: [64, 64] \\
			& & & Attention: $d_k=64$\\
			& & & Decoder: [64, 64]\\
			Number of parameters & 3.0e4 & ? & 3.4e4 \\
			Variable input size & No & No & Yes \\
			Permutation invariant & No & Yes & Yes \\
			\bottomrule
		\end{tabular}
	\end{threeparttable}
\end{table}

>>>>

% Custom packages
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{array, makecell}
\usepackage{tikz}
\usepackage{xspace}
\usepackage{float}
\usetikzlibrary{arrows,automata}
\usepackage{pgfplots}
\usetikzlibrary{intersections}
\usepackage{subcaption}
\usepackage[flushleft]{threeparttable}
\usepackage{cases}
\usepackage{multirow}

@@

This observation has led the community to turn to learning-based methods, which bear the promise of leveraging data to automatically learn a complex driving policy. In the imitation learning approach, a policy can be trained in a supervised manner to imitate human driving decisions \citep[e.g.][]{Pomerleau1989, Ross2011, Bojarski2016, Xu2016, Eraqi2017, Codevilla2017, Rehder2017c, Rezagholiradeh2018, Rhinehart2018, Bansal2018, Rhinehart2019}. Because the cost of human driving data collection at large scale can be prohibitive, another promising approach is train a policy in simulation using reinforcement learning \citep[e.g.][]{Cardamone2009, Ross2011, Mukadam2017, Chen2017, Isele_2018, Ha2018, Kendall2019}.

@@

Our contributions are the following: first, we propose an attention-based architecture for decision-making involving social interactions. This architecture allows to satisfy the variable-size and permutation invariance requirements even when using a \emph{list of features} representation. It also naturally accounts for interactions between the ego-vehicle and any other traffic participant.
Second, we evaluate our model on a challenging intersection-crossing task involving up to 15 vehicles perceived simultaneously. We show that our proposed method provides significant quantitative improvements, and that it enables to capture interaction patterns in a way that is visually interpretable.

@@

In order to apply a reinforcement learning algorithm such as DQN to an autonomous driving problem, a state space $S$ must first be chosen, that is, a representation of the scene. When social interactions are relevant to the decision, the state should at leat compose a description of every nearby vehicle. A vehicle driving on a road can be described in the most general way by it's continuous position, heading and velocity. Then, the joint state of a road traffic with one ego-vehicle denoted $s_0$ and $N$ other vehicles can be described by the (unordered) set of individual vehicle states:

\begin{equation*}
s = \left\lbrace s_i \right\rbrace_{i \in [0, N]}\qquad
\text{where}\qquad
s_i = \begin{bmatrix}
x_i & y_i & v^x_i & v^y_i & \cos\psi_i & \sin \psi_i
\end{bmatrix}^T
\end{equation*}

@@

This representation, that we call \emph{list of features}, is illustrated in \autoref{fig:representation} (left) and was used for instance in \citep{Bai2015, Gindele2015, Song2016, Sunberg2017, Paxton2017, Galceran2017, Chen2017}.

@@

A popular way to address this limitations is to use a \emph{spatial grid} representation. Instead of explicitly representing spatial information as variables $x, y$ along with other features $f$ directly inside the state $\{s_i=(x_i,y_i,f_i)\}_{i\in[0,N]}$ indexed on the vehicles, they are instead represented implicitly through the layout of several feature variables $f_{ij}$ organized in a tensor structure, where the $(i,j)$ indexes refer to the cells of a quantization of the 2D-space. This representation is illustrated in \autoref{fig:representation} (right). Note that the size of this tensor is related to the area covered divided by the length of the quantization, which reflects a trade-off between accuracy and dimensionality.
In an occupancy grid, the $f$ features contains presence information (0-1) and additional channels such as velocity and heading, as in \citep[e.g.][]{Isele2017, Fridman2018, Bansal2018, Rehder2017c}. Another example is the use of top-view RGB images \citep[e.g.][]{Bagnell2010, Rehder2017, Rehder2017c, Liu2018}.

This permutation invariance property can also be implemented within the policy architecture. A general technique to achieve this is to treat each entity similarly in the early stages -- e.g. through weight sharing -- before reducing them with a projection operator that is itself invariant to permutations, for instance a max-pooling as in \citep{Chen2017} or an average as in \citep{Qi2016}. A particular instantiation of this idea is attention mechanisms.

\paragraph{Attention mechanisms} {The attention architecture was introduced to enable neural networks to discover inter-dependencies within a variable number of inputs.
It has been used for pedestrian trajectory forecasting in~\cite{Vemula2018} with spatiotemporal graphs and in~\cite{Sadeghian2019CVPR} with spatial and social attention using a generative neural network. In~\cite{Sadeghian2018ECCV}, attention over top-view road scene images for car trajectory forecasting is used. Multi-head attention mechanism has been developed in~\cite{Vaswani2017} for sentence translation. In~\cite{Messaoud2019} a mechanism called non-local multi-head attention is developed. However, this is a spatial attention that does not allow vehicle-to-vehicle attention. In the present work, we use a multi-head social attention mechanism to capture vehicle-to-ego dependencies and build varying input size and permutation invariance into the policy model.}

@@

Out of a complex scene description, the model should be able to filter information and consider only what is relevant for decision. In other words, the agent should \emph{pay attention} to vehicles that are close or conflict with the planned route.

The proposed architecture is presented in \autoref{fig:architecture}. It is used to represent the $Q$-function that will be optimized by the DQN algorithm. It is composed of a first linear encoding layer whose weights are shared between all vehicles. At that point, the embeddings only contain individual features of size $d_x$. They are then fed to an ego-attention layer, composed of several heads stacked together. Such an attention head is illustrated in \autoref{fig:ego-attention}, and works in the following way: in order to select a subset of vehicles depending on the context, the ego-vehicle  first emits a query $Q = [q_0]\in\Real^{1 \times d_k}$, computed with a linear projection $L_q\in\Real^{d_x \times d_k}$ of its embedding. This query is then compared to a set of keys $K = [k_0, \dots, k_N]\in\Real^{N \times d_k}$ containing descriptive features $k_i$ for each vehicle, again computed with a shared linear projection $L_k\in\Real^{d_x \times d_k}$. The similarity between the query $q_0$ and any key $k_i$ is assessed by their dot product $q_0 k_i^T$. These similarities are then scaled by the inverse-square-root-dimension $1/\sqrt{d_k}$\footnote{This scaling is due to the fact that the dot-product of two independent random vectors with mean 0,  variance 1, and dimension $d_k$, is a random variable with mean 0 and variance $d_k$} and normalized with a softmax function $\sigma$ across vehicles. We obtain a stochastic matrix called the \emph{attention matrix}, which is finally used to gather a set of output value $V = [v_0, \dots, v_N]$, where each value $v_i$ is a feature computed with a shared linear projection $L_v\in\Real^{d_x \times d_v}$. Overall, the ego-attention computation for each head can be written as:
\begin{equation}
\text{output}=\underbrace{\sigma\left(\frac{QK^T}{\sqrt{d_k}}\right)}_{\text{attention matrix}}V
\label{eq:selfattention}
\end{equation}
The outputs from all heads are finally combined with a linear layer, and the resulting tensor is then added to the ego encoding as in residual networks. We can easily see that the result is permutation invariant: indeed, a permutation $\tau$ will change the order of the rows in keys $K$ and values $V$ in \eqref{eq:selfattention} but will keep their correspondence. The final result is a dot product of values and key-similarities, which is independent of the ordering.

\begin{figure}[tp]
	\centering
	\begin{tikzpicture}
	\node(X1){ego};
	\node[below of=X1, node distance=0.7cm](X2){vehicle$_{1}$};
	\node[below of=X2, node distance=0.6cm](X3){$\vdots$};
	\node[below of=X3, node distance=0.7cm](X4){vehicle$_{N}$};

@@

	\node[draw, rectangle, right of=ENC1, node distance=2.0cm, below=-0.4cm, fill=white](TRANS3){\rotatebox{90}{ Ego-attention }};
	\node[draw, rectangle, right of=ENC1, node distance=1.9cm, below=-0.3cm, fill=white](TRANS2){\rotatebox{90}{ Ego-attention }};
	\node[draw, rectangle, right of=ENC1, node distance=1.8cm, below=-0.2cm, fill=white](TRANS1){\rotatebox{90}{ Ego-attention }};

@@

	\draw (DEC1.east) -- (Y1.west);
	\end{tikzpicture}
	\caption{Block diagram of our model architecture.}
	\label{fig:architecture}

@@

\begin{figure}[tp]
	\centering
	\begin{tikzpicture}[scale=1, every node/.style={scale=1}]
	\node(X1){ego encoding};
	\node[below of=X1, node distance=2cm](X2){$\vdots$};
	\node[below of=X2, node distance=2cm](X3){vehicle$_{N}$ encoding};

@@

@@

\begin{table}[tp]
	\centering
	\begin{threeparttable}
		\caption{Characteristics of the agents}
		\label{tab:agents}
		\begin{tabular}{lccc}
			\toprule
			Architecture & \MLPC & \MLPG & \EgoAtt \\
			\midrule 
			Input sizes & [15, 7] & [?, ?, 7] & [1-15, 7] \\
			Layers sizes & [128, 128] &  [64, 64] & \makecell[l]{Encoder: [64, 64] \\Attention: 2 heads, $d=32$ \\ Decoder: [64, 64]} \\
			Number of parameters & 3.0e4 & ? & 3.4e4 \\
			Variable input size & No & No & Yes  \\
			Permutation invariant & No & Yes & Yes \\
			\bottomrule
		\end{tabular}
	\end{threeparttable}
\end{table}
